Automatic Extraction of JavaScript Formal Semantics from ECMAScript Specifications

### 2019-08-08 icse/compiler.tex
Written overview

Abstract algorithms in ECMAScript specifications are written in structured
natural languages in HTML files. An algorithm consists of ordered steps
and it might contain sub-steps as well. For example,
the \( \code{ToPrimitive} \) algorithm in Figure~\ref{fig:to-primitive}
has three steps and second step has seven sub-steps.
Moreover, the tokens of each step has its own HTML tag and each tag
has the following meaning:
\[
  \begin{array}{c|l}
    \text{HTML tags} & \text{meanings}\\\hline\hline
    \code{<var>} & \text{variables}\\\hline
    \code{<emu-grammar>} & \text{productions}\\\hline
    \code{<emu-nt>} & \text{non-terminal syntax}\\\hline
    \code{<code>} & \text{ECMAScript codes}\\\hline
    \code{<emu-const>} & \text{constant values}\\\hline
    \code{<emu-val>} & \text{values}\\\hline
    \code{<ol>} & \text{ordered sub-steps}\\\hline
    \code{<ul>} & \text{unordered sub-steps}\\\hline
    \code{<sup>} & \text{superscripts}\\\hline
    \text{otheriwse} & \text{simple texts}\\\hline
  \end{array}
\]
We try to keep the tag information for each token to generate more
precise conversion rules. For example, if and only if a token has a tag
\( \code{<var>} \), it is a parameter or a local variable.
Thus, it is possible to construct a conversion rule precisly discriminates
identifiers and other components.

@@

\item \textbf{Primitive Values} \( \ires \) supports ECMAScript primitive values except
the symbols that is newly introduced after ECMAScript 6. Thus, only five ECMAScript primitives
are defined in \( \ires \); numbers, strings, booleans, undefined, and null.
Moreover, \( \ires \) also supports the unique \( \code{absent} \) value to represent
the absence of parameters. For example, the second parameter \( \code{PreferredType} \)
of \( \code{ToPrimitive} \) algorithm in Figure~\ref{fig:to-primitive} might be
not present in some function calls. Then, the parameter has the \( \code{absent} \) value
to represent its absence.

@@

To generate the corresponding \( \ires \) function from syntactic parse tree,
we describe the mapping from each parsing rule into the corresponding \( \ires \) components.
For example, we describe the following mapping for the previous example parsing rules:
\begin{lstlisting}[style=myScalastyle]
// statements
val Stmt = ... ^^ { case _ ~ x ~ _ ~ y => ESLet(x, y) }
// expressions
val Expr = (
  // identifiers
  ... ^^ { x => x } |
  // return if abrupt
  ... ^^ { _ ~ e => ESReturnIfAbrupt(e) } |
  // function calls
  ... ^^ { case x ~ _ ~ y ~ _ => ESCall(x, y) } |
  // lists
  ... ^^ { case _ ~ x ~ _ => ESList(e) }
)
\end{lstlisting}
Finally, sub-step \( i \) in the \( \code{ToPrimitive} \) is converted
into the following instructions:
\begin{lstlisting}[style=ires]
let result = ? (Call exoticToPrim input (new [hint]))
\end{lstlisting}

>>>>

Abstract algorithms in ECMAScript specifications are written in structured
natural languages in HTML files. An algorithm consists of ordered steps
and it might contain sub-steps as well. For example,
the \( \code{ToPrimitive} \) algorithm in Figure~\ref{fig:to-primitive-es}
has three steps and second step has seven sub-steps.
Moreover, the tokens of each step has its own HTML tag and each tag
has the following meaning:
\[
  \begin{array}{c|l}
    \text{HTML tags} & \text{meanings}\\\hline\hline
    \code{<var>} & \text{variables}\\\hline
    \code{<emu-grammar>} & \text{productions}\\\hline
    \code{<emu-nt>} & \text{non-terminal syntax}\\\hline
    \code{<code>} & \text{ECMAScript codes}\\\hline
    \code{<emu-const>} & \text{constant values}\\\hline
    \code{<emu-val>} & \text{values}\\\hline
    \code{<ol>} & \text{ordered sub-steps}\\\hline
    \code{<ul>} & \text{unordered sub-steps}\\\hline
    \code{<sup>} & \text{superscripts}\\\hline
    \text{otheriwse} & \text{simple texts}\\\hline
  \end{array}
\]
We try to keep the tag information for each token to generate more
precise conversion rules. For example, if and only if a token has a tag
\( \code{<var>} \), it is a parameter or a local variable.
Thus, it is possible to construct a conversion rule precisly discriminates
identifiers and other components.

@@

\item \textbf{Primitive Values} \( \ires \) supports ECMAScript primitive values except
the symbols that is newly introduced after ECMAScript 6. Thus, only five ECMAScript primitives
are defined in \( \ires \); numbers, strings, booleans, undefined, and null.
Moreover, \( \ires \) also supports the unique \( \code{absent} \) value to represent
the absence of parameters. For example, the second parameter \( \code{PreferredType} \)
of \( \code{ToPrimitive} \) algorithm in Figure~\ref{fig:to-primitive-es} might be
not present in some function calls. Then, the parameter has the \( \code{absent} \) value
to represent its absence.

@@

To generate the corresponding \( \ires \) function from syntactic parse tree,
we describe the mapping from each parsing rule into the corresponding \( \ires \) components.
For example, we describe the following mapping for the previous example parsing rules:
\begin{lstlisting}[style=myScalastyle]
// statements
val Stmt = ... ^^ { case _ ~ x ~ _ ~ y => ESLet(x, y) }
// expressions
val Expr = (
  // identifiers
  ... ^^ { x => x } |
  // return if abrupt
  ... ^^ { _ ~ e => ESReturnIfAbrupt(e) } |
  // function calls
  ... ^^ { case x ~ _ ~ y ~ _ => ESCall(x, y) } |
  // lists
  ... ^^ { case _ ~ x ~ _ => ESList(e) }
)
\end{lstlisting}
Finally, sub-step \( i \) in the \( \code{ToPrimitive} \) is converted
into the following instruction:
\begin{lstlisting}[style=ires]
let result = ? (Call exoticToPrim input (new [hint]))
\end{lstlisting}

### 2019-08-08 icse/intro.tex
Written overview

JavaScript is one of widely used programming languages not only for client-sides
but also for server-side programming~\cite{???} and even operating systems~\cite{???}.
However, developers suffer from the intricate semantics of JavaScript and
such complexities cause security vulnerabilities. For example, the following JavaScript
code seems to be tautology:
\begin{lstlisting}[style=myJSstyle]
function f(x) { return x == !x; }
\end{lstlisting}
Unfortunately, it returns \( \code{false} \) when the argument is an empty array
\( \code{[]} \). To correctly understand such situations, the understanding of
formal semantics of JavaScript is necessary.

@@

Moreover, it is not suitable approach to cope with the updates of ECMAScript specifications.
Most of full formal semantics of ECMAScript only supports ECMAScript 5.1 that was
released in December 2009. Until 2014 years, it was not big deal because ECMAScript
specifications are rarely updated. However, in late 2014, The Ecma Technical
Committee 39 (TC39) decided to release ECMAScript on a yearly cadence. From the
ECMAScript 6, officially ECMAScript 2015, they updated specifications every years
in June. Now, no researches correctly deal with recent ECMAScript features;
lexical bindings (\( \code{let} \)), spread operator (\( \code{...} \),
classes, for-of operators, async functions, or generators.

To alleviate such problems, we propose \textit{Automatic Semantics Extractor (ASE)}
for ECMAScript specifications. The tool automatically generates JavaScript parser
from syntax provided by ECMAScript specifications. It also extracts formal
JavaScript semantics in semi-automatic ways. It is semi-automatic because
it depends on manually defined \textit{conversion rules}. Each rule describes how abstract
algorithms in specifications are converted into functions of our core language \( \ires \).
However, to save the labor of defininig such rules, we also propose \textit{rule generation assistant}
that suggests new conversion rules based on not yet convertable algorithms.
The assistant help us deal with changed or newly introduced algorithms defined in new version of
ECMAScript specifications as well.

>>>>

JavaScript is one of widely used programming languages not only for client-sides
but also for server-side programming~\cite{???} and even operating systems~\cite{???}.
However, developers suffer from the intricate semantics of JavaScript and
such complexities cause security vulnerabilities. For example, the following JavaScript
code seems to always return \( \code{false} \):
\begin{lstlisting}[style=myJSstyle]
function f(x) { return x == !x; }
\end{lstlisting}
Unfortunately, it returns \( \code{true} \) when the argument is an empty array
\( \code{[]} \). To correctly understand such situations, the understanding of
formal semantics of JavaScript is necessary.

@@

Moreover, it is not suitable approach to cope with the updates of ECMAScript specifications.
Most of full formal semantics of ECMAScript only supports ECMAScript 5.1 that was
released in December 2009. Until 2014, it was not big deal because ECMAScript
specifications are rarely updated. However, in late 2014, The Ecma Technical
Committee 39 (TC39) decided to release ECMAScript on a yearly cadence. From the
ECMAScript 6, officially ECMAScript 2015, they updated specifications every years
in June. Now, no researches correctly deal with recent ECMAScript features;
lexical bindings (\( \code{let} \)), spread operator (\( \code{...} \)),
classes, for-of operators, async functions, or generators.

To alleviate such problems, we propose \textit{Automatic Semantics Extractor (ASE)}
for ECMAScript specifications. The tool automatically generates JavaScript parser
from syntax provided by ECMAScript specifications. It also extracts formal
JavaScript semantics in semi-automatic ways. It is semi-automatic because
it depends on manually defined \textit{conversion rules}. Each rule describes how abstract
algorithms in specifications are converted into functions of our core language \( \ires \).
However, to save the labor of defininig such rules, we also propose \textit{rule generation assistant}.
It suggests new conversion rules based on statistical analysis of not yet convertable algorithms.
The assistant helps us deal with changed or newly introduced algorithms defined in new versions of
ECMAScript specifications as well.

### 2019-08-08 icse/overview.tex
Written overview

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{img/overview.png}
  \caption{Overall structure of Automatic Semantics Extractor (ASE)}
  \label{fig:overview}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[height=18em]{img/to_primitive.png}
  \begin{lstlisting}[style=ires]
"ToPrimitive" (input, PreferredType) => {
  if (= (Type input) "Object") {
    if (= PreferredType absent) let hint = "default"
    else if (= PreferredType "String") let hint = "string"
    else let hint = "number"
    let exoticToPrim = ? (GetMethod input SYMBOL_toPrimitive)
    if (! (= exoticToPrim undefined)) {
      let result = ? (Call exoticToPrim input (new [hint]))
      if (! (= (Type result) "Object")) return result
      return (Throw INTRINSIC_TypeErrorPrototype)
    }
    if (= hint "default") hint = "number"
    return ? (OrdinaryToPrimitive input hint)
  }
  return input
}
  \end{lstlisting}
  \caption{\( \code{ToPrimitive} \) abstract algorithms
  and its generated core program.}
  \label{fig:to-primitive}
\end{figure}

\begin{itemize}
  \item Sytnax: automatically generate parser from ES-BNF
    \begin{itemize}
      \item ES-BNF features
      \item Packrat Parsers with 1-lookahead
      \item Example
    \end{itemize}
  \item Semantics: automatically convert abstract algorithms into Core programs
    \begin{itemize}
      \item Abstract algorithms
      \item Example
    \end{itemize}
\end{itemize}
\begin{itemize}
  \item Figure of overall structure
  \item Explanation of each module
\end{itemize}

>>>>

In this section, we introduce the overall structure of ASE and the outline of
this paper. Figure~\ref{fig:overview} depicts the overall structure of ASE.
It mechanizes to extract syntax and semantics of ECMAScript using
two main module; the \textit{parser generator} and the \textit{algorithm compiler}.
If first extracts information via the spec extractor.
ECMAScript specifications are provided as a readable HTML file.
Thus, we develop the spec extractor in JavaScript to automatically
extract their contents.

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{img/overview.png}
  \caption{Overall structure of Automatic Semantics Extractor (ASE)
  for ECMAScript specifications}
  \label{fig:overview}
\end{figure}

ECMAScript specifications describe their lexical and syntactic grammars
based on their own extension of Backus-Naur Form (BNF).
We call it as \( \bnfes \), BNF for ECMAScript specifications.
It supports additional notations for parametric productions, conditional
right-hand sides, optional symbols, positive/negative lookaheads,
exclusive symbols, and no line-terminator symbols. We first formally define
\( \bnfes \) and build the spec extractor to extract grammars written in \( \bnfes \).
Then, we carefully construct the \textit{parser generator} for them by extending
the Scala parser combinators. Moreover, we propose \textit{lookahead parsers}
to parse \( \bnfes \) for correct and fast parsing.
A lookahead parser keeps track of its lookahead, which is the set of
possible next tokens. With lookahead parsers,
the parser generator automatically generate parsers for ECMAScript grammars.
The generated parsers are readable and one-to-one corresponds to each
grammar components.
For example, Firugre~\ref{fig:array-literal-es} is ArrayLiteral production in
ECMAScript 2019 specification and Figure~\ref{fig:array-literal-parser}
is the generated parser of the algorithm written in Scala.
In the first right-hand side, the optional \( Elision \) token
is represented as \( \code{opt(Elision)} \).
The parametric non-terminal \( ElementList \) with arguments \( Yield \)
and \( Await \) is represented as \( \code{ElementList(Yield, Await)} \).

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \includegraphics[width=\textwidth]{img/array_literal.png}
    \caption{ArrayLiteral production in ECMAScript 2019 specification.}
    \label{fig:array-literal-es}
  \end{subfigure}
  \begin{subfigure}[t]{0.5\textwidth}
    \begin{lstlisting}[style=myScalastyle]
type P[T] = List[Boolean] => LAParser[T]
lazy val ArrayLiteral: P[ArrayLiteral] = memo {
  case List(Yield, Await) =>
  "[" ~ opt(Elision) ~ "]" ^^ ArrayLiteral0 |
  "[" ~ ElementList(Yield, Await) ~ "]" ^^ ArrayLiteral1 |
  "[" ~ ElementList(Yield, Await) ~ "," ~ opt(Elision) ~ "]" ^^ ArrayLiteral2
}
    \end{lstlisting}
    \caption{The generated parser for ArrayLiteral production.}
    \label{fig:array-literal-parser}
  \end{subfigure}
  \caption{ArrayLiteral grammar production in ECMAScript 2019.}
  \label{fig:array-literal}
\end{figure}

The semantics in ECMAScript specifications are described as
abstract algorithms. They are written in structured natural languages
with ordered steps and tagged tokens. For example, Figure~\ref{fig:to-primitive-es}
is an example abstract algorithm for \( \code{ToPrimitive} \) in
ECMAScript 2019. The spec extractor extracts each abstract algorithm
with HTML tags because we utilize tag information to discriminate
different kinds of tokens. Then, the \textit{algorithm compiler} converts
given abstract algorithms into funcitons of our core language \( \ires \).
The compilation is dependent on manually defined conversion rules
that consists of two parts; parsing rules and mapping from each rule into \( \ires \)
component. We carefully define \( \ires \) to directly match instructions with
the corresponding algorithm steps. For example, Figure~\ref{fig:to-primitive-ires}
describes the generated \( \ires \) function for the \( \code{ToPrimitive} \)
algorithm. Each instruction of the generated \( \ires \) function
is easily matched with the corresponding algorithm step
and also they have similar silhouettes.

\begin{figure*}[t]
  \centering
  \begin{subfigure}{0.4\textwidth}
    \includegraphics[width=\textwidth]{img/to_primitive.png}
    \subcaption{\( \code{ToPrimitive} \) abstract algorithm in ES2019.}
    \label{fig:to-primitive-es}
  \end{subfigure}
  \qquad
  \begin{subfigure}{0.48\textwidth}
    \begin{lstlisting}[style=ires]
"ToPrimitive" (input, PreferredType) => {
  if (= (Type input) "Object") {
    if (= PreferredType absent) let hint = "default"
    else if (= PreferredType "String") let hint = "string"
    else let hint = "number"
    let exoticToPrim = ? (GetMethod input SYMBOL_toPrimitive)
    if (! (= exoticToPrim undefined)) {
      let result = ? (Call exoticToPrim input (new [hint]))
      if (! (= (Type result) "Object")) return result
      return (Throw INTRINSIC_TypeErrorPrototype)
    }
    if (= hint "default") hint = "number"
    return ? (OrdinaryToPrimitive input hint)
  }
  return input
}
    \end{lstlisting}
    \subcaption{The \( \ires \) funciton for \( \code{ToPrimitive} \)
    abstract algorithm.}
    \label{fig:to-primitive-ires}
  \end{subfigure}
  \caption{\( \code{ToPrimitive} \) abstract algorithms
  and its generated core program.}
  \label{fig:to-primitive}
\end{figure*}

Finally, the generated \( \ires \) functions with mannual initial settings
are applied into the \( \ires \) interpreter. We could parse a
given JavaScript program using the generated JavaScript parser
to get the abstract syntax tree (AST). Then, the \( \ires \) interpreter
evaluates the given AST and provides evaluation results.

From now, we describe the detailed techniques of ASE.
Section 3 formallly defines \( \bnfes \) and how to automatically
generate parsers from given \( \bnfes \) using lookahead parsers.
Section 4 explains abstract algorithms with tagged tokens
and how to compile them into \( \ires \) functions based on
given conversion rules. Section 5 describes the implemented details
about initial settings to evaluate JavaScript programs. Section 6
discusses the effectiveness of ASE with four research questions.
Section 7 discusses related work and Section 8 concludes.

### 2019-08-08 icse/parser.tex
Written overview

@@

\begin{figure}
  \centering
  \includegraphics[height=6em]{img/array_literal.png}
  \[
    \begin{array}{l}
      \NT{ArrayLiteral}(Yield, Await) ::=\\
      \qquad \T{[} \; Elision? \; \T{]}\\
      \qquad \T{[} \; ElementList(Yield, Await) \; \T{]}\\
      \qquad \T{[} \; ElementList(Yield, Await) \; \T{,} Elision? \; \T{]}\\
    \end{array}
  \]
  \caption{Array literals defined in ES2019 and its \( \bnfes \)}
  \label{fig:array-literal}
\end{figure}

@@

\subsubsection{Parser generations}\label{sec:convert-bnfes}
The next step is to automatically extract parsers from the given \( \bnfes \) grammar.
The conversion from \( \bnfes \) symbols into Scala codes is defined as follows:
\[
  \begin{array}{c|c}
    \bnfes \; \text{symbols} & \text{Scala codes}\\\hline\hline
    \epsilon & \code{MATCH}\\\hline
    \T{a} & \code{"a"}\\\hline
    \NT{A}(\argument_1, \argument_n) & \code{A(a1, .. , an)}\\\hline
    \symb? & \code{opt(s)}\\\hline
    \symb \butnot \symb' & \code{s\textbackslash s'}\\\hline
    \nolt & \code{NoLineTerminator}\\\hline
  \end{array}
\]
The \( \code{MATCH} \) denotes the empty sequence of lookahead parsers.
Each String literals are implicitly converted into lookahead parsers
through Scala implicit conversions. The \( \code{opt(s)} \) funciton
is same with \( \code{s | MATCH} \). We also define \( \code{\textbackslash} \)
opertor between parsers to support exclusive parsers.
Finally, we support the \( \code{NoLineTerminator} \) parser.
It is a little bit tricky because it should hook the whitespace parsers
and check whether it contains line terminators.
However, it is possible in our approach because
we also automatically generates lexers not only parsers of ECMAScript syntax.
For the final result, the automatically synthesized parser from the
production \( ArrayLiteral \) in Figure~\ref{fig:array-literal} is defined as follows:
\begin{lstlisting}[style=myScalastyle]
lazy val ArrayLiteral: List[Boolean] => LAParser[ArrayLiteral] = memo {
  case args @ List(Yield, Await) =>
    ((MATCH <~ "[") ~ opt(Elision) <~ "]") ^^ {
      case _ ~ x0 => ArrayLiteral0(x0)
    } | ((MATCH <~ "[") ~ ElementList(Yield, Await) <~ "]") ^^ {
      case _ ~ x0 => ArrayLiteral1(x0)
    } | (((MATCH <~ "[") ~ ElementList(Yield, Await) <~ ",") ~ opt(Elision) <~ "]") ^^ {
      case _ ~ x0 ~ x1 => ArrayLiteral2(x0, x1)
    }
}
\end{lstlisting}

>>>>

\begin{figure}
  \centering
  \[
    \begin{array}{l}
      \NT{ArrayLiteral}(Yield, Await) ::=\\
      \qquad \T{[} \; Elision? \; \T{]}\\
      \qquad \T{[} \; ElementList(Yield, Await) \; \T{]}\\
      \qquad \T{[} \; ElementList(Yield, Await) \; \T{,} Elision? \; \T{]}\\
    \end{array}
  \]
  \caption{ArrayLiteral production in \( \bnfes \).}
  \label{fig:array-literal-bnfes}
\end{figure}

For example,

@@

@@

\subsubsection{Parser generations}\label{sec:convert-bnfes}
The next step is to automatically extract parsers from the given \( \bnfes \) grammar.
The conversion from \( \bnfes \) symbols into Scala codes is defined as follows:
\[
  \begin{array}{c|c}
    \bnfes \; \text{symbols} & \text{Scala codes}\\\hline\hline
    \epsilon & \code{MATCH}\\\hline
    \T{a} & \code{"a"}\\\hline
    \NT{A}(\argument_1, \argument_n) & \code{A(a1, .. , an)}\\\hline
    \symb? & \code{opt(s)}\\\hline
    \symb \butnot \symb' & \code{s\textbackslash s'}\\\hline
    \nolt & \code{NoLineTerminator}\\\hline
  \end{array}
\]
The \( \code{MATCH} \) denotes the empty sequence of lookahead parsers.
Each String literals are implicitly converted into lookahead parsers
through Scala implicit conversions. The \( \code{opt(s)} \) funciton
is same with \( \code{s | MATCH} \). We also define \( \code{\textbackslash} \)
opertor between parsers to support exclusive parsers.
Finally, we support the \( \code{NoLineTerminator} \) parser.
It is a little bit tricky because it should hook the whitespace parsers
and check whether it contains line terminators.
However, it is possible in our approach because
we also automatically generates lexers not only parsers of ECMAScript syntax.
For the final result, the automatically synthesized parser from the
production \( ArrayLiteral \) in Figure~\ref{fig:array-literal} is defined
as the parsers defined in Figure~\ref{fig:array-literal}.
